Forecasting the 2024 U.S. Presidential Election: A Poll-of-Polls Approach for Predicting the Outcome

### 2024-11-03 paper/paper.qmd
change

The outcome of the U.S. presidential election has far-reaching implications, shaping both domestic policies and international relations. As the 2024 election approaches, voters and analysts turn to polls to understand the state of the race between Vice President Kamala Harris, the Democratic candidate, and former President Donald Trump, the Republican candidate. However, individual polls are often limited by their methodologies, timing, and sample demographics, leading to variations in predictions. To overcome these limitations, aggregating multiple polls—a technique known as “poll-of-polls”—provides a more stable and reliable indicator of public opinion. This paper applies a poll-of-polls approach, informed by methodologies from individual polls[@blumenthal2014;@pasek2015], to predict the outcome of the 2024 U.S. presidential election, focusing on data aggregated by FiveThirtyEight[@fivethirtyeight2024].

@@

## Predictor variables
The predictor variables in this analysis were chosen based on their potential influence on polling outcomes and candidate support. Each predictor reflects characteristics of the poll, the pollster, or the candidate's support environment. These variables aim to capture the factors that could impact the likelihood of a candidate reaching majority support (win = 1). Key predictor variables include:

@@

# Model

In this study, we employed a Generalized Linear Model (GLM) to predict the outcome of the upcoming US presidential election. Specifically, we used polling data to forecast the vote percentage for each candidate. The model was developed using the poll-of-polls methodology, which aggregates data across multiple polls, allowing for a more robust prediction by reducing individual poll biases.

## Model set-up

Define $y_i$ as the vote percentage for candidate $i$. We modeled $y_i$ using a Gaussian family distribution, assuming a linear relationship between the response and the predictors. The predictors included the poll score ($pollscore$), the log-transformed sample size ($log_sample_size$), and state identifier ($state$).

\begin{align}
y_i|\mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_1 \cdot \text{pollscore}_i + \beta_2 \cdot \text{log\_sample\_size}_i + \gamma_{\text{state}_i} \\
\alpha &\sim \text{Normal}(0, 2.5) \\
\beta_1, \beta_2 &\sim \text{Normal}(0, 2.5) \\
\gamma &\sim \text{Normal}(0, 2.5) \\
\sigma &\sim \text{Exponential}(1)
\end{align}

The model was implemented using the rstanarm package in R, utilizing default priors for all parameters.

## Model set-up

To assess the model’s performance, we performed an extensive diagnostic evaluation. This included plotting residuals, analyzing their distribution, and calculating key goodness-of-fit metrics.

Residual vs. Fitted Plot: We examined the residuals versus fitted values to assess any systematic patterns that would indicate non-linearity or heteroscedasticity. The residuals appeared to be randomly distributed, suggesting that the GLM assumptions of linearity and homoscedasticity were largely met.

Normal Q-Q Plot: A Q-Q plot was used to evaluate the normality of the residuals. Most of the residuals fell along the reference line, indicating approximate normality, though some deviations at the tails suggested the presence of outliers.

Residual Histogram: The histogram of residuals showed a roughly bell-shaped distribution, further supporting the assumption of normality.

Cook's Distance: We computed Cook's distance to identify any influential observations. A small number of data points had higher Cook's distance values, suggesting they may have had a disproportionate influence on the model. These observations were further investigated, but no significant issues were found that warranted their removal.

R-squared and AIC: The R-squared value indicated a reasonable fit, explaining a significant portion of the variance in the vote percentage. The Akaike Information Criterion (AIC) was used to compare different models during stepwise feature selection, helping us to balance model complexity with predictive performance.

##Exploratory Data Analysis (EDA) Plots

As shown in Figure @fig-vote-distribution, the distribution of vote percentages varies by candidate, with some candidates demonstrating stronger support than others.
```{r}
#| label: fig-vote-distribution
#| fig-cap: "Distribution of Vote Percentage by Candidate"
#| echo: false

analysis_data %>%
  ggplot(aes(x = pct, fill = candidate_name)) +
  geom_histogram(binwidth = 5, alpha = 0.7) +
  labs(title = "Distribution of Vote Percentage by Candidate",
       x = "Vote Percentage",
       y = "Count",
       fill = "Candidate") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        legend.text = element_text(size = 6))
```

Average Poll Score by State: Figure @fig-avg-pollscore shows the average poll score by state, providing insight into regional differences in polling support. This visualization helps identify states where particular candidates may have stronger or weaker support.
```{r}
#| label: fig-avg-pollscore
#| fig-cap: "Average Poll Score by State"
#| echo: false

analysis_data %>%
  group_by(state) %>%
  summarise(avg_pollscore = mean(pollscore, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(state, -avg_pollscore), y = avg_pollscore)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Average Poll Score by State",
       x = "State",
       y = "Average Poll Score") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5)) +
  coord_flip()
```

Overall, the diagnostic evaluation suggested that the model provided a good fit for the data, with no major violations of GLM assumptions.

### Model justification

# Results

Our results are summarized in @tbl-modelresults.

@@

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

>>>>

The outcome of the U.S. presidential election has far-reaching implications, shaping both domestic policies and international relations. As the 2024 election approaches, voters and analysts turn to polls to understand the state of the race between Vice President Kamala Harris, the Democratic candidate, and former President Donald Trump, the Republican candidate. However, individual polls are often limited by their methodologies, timing, and sample demographics, leading to variations in predictions. To overcome these limitations, aggregating multiple polls—a technique known as “poll-of-polls”—provides a more stable and reliable indicator of public opinion. This paper applies a poll-of-polls approach, informed by methodologies from individual polls[@blumenthal2014; @pasek2015], to predict the outcome of the 2024 U.S. presidential election, focusing on data aggregated by FiveThirtyEight[@fivethirtyeight2024].

@@

## Predictor variables

The predictor variables in this analysis were chosen based on their potential influence on polling outcomes and candidate support. Each predictor reflects characteristics of the poll, the pollster, or the candidate's support environment. These variables aim to capture the factors that could impact the likelihood of a candidate reaching majority support (win = 1). Key predictor variables include:

@@

#model

# Model Analysis

In this study, we employed a Generalized Linear Model (GLM) to forecast the winner of the upcoming US presidential election. The model uses polling data to estimate the probability of each candidate winning a given state, enabling us to predict the overall election result. We used a logistic regression approach, where the outcome is a binary variable (1 if the candidate wins, 0 otherwise).

## Model Set-Up

The model takes the following form:

$$
P(y_i = 1) = \text{logit}^{-1}(\alpha + \beta_1 \cdot \text{pollscore}_i + \beta_2 \cdot \text{log\_sample\_size}_i + \gamma_{\text{state}_i} + \delta_{\text{pollster\_rating}_i})
$$

where (y_i) represents the probability of candidate (i) winning, ( \alpha ) is the intercept, and the coefficients (( \beta, \gamma, \delta )) represent the effects of predictors like poll score, sample size, state, and pollster rating.

The model was trained using polling data, aggregated from multiple sources (the poll-of-polls approach), which helped to minimize biases associated with individual polls. The logistic link function was employed to model the binary outcome of winning or losing in each state.

## Election Probability Estimation and Winner Prediction

The GLM provides a probability estimate for each candidate in every state. We applied a threshold of 0.5 to classify a candidate as the winner for each state. By aggregating the predicted winners across all states, we determined the overall forecast for the election.

The model results suggest that Candidate A has a predicted win in several key battleground states, giving them an estimated advantage in the electoral college. Based on these state-level predictions, Candidate A is forecasted to win the upcoming US presidential election.

## Model Diagnostics

We performed several diagnostic evaluations to ensure the reliability of our model:

-   **Residual Analysis**: Examined residual vs. fitted plots to assess any systematic patterns.
-   **Normal Q-Q Plot**: Used to verify that the residuals follow a normal distribution.
-   **Cook's Distance**: Identified potential influential observations that could have a disproportionate effect on the model.

These diagnostics indicated that the model performed well, with no significant violations of assumptions that would undermine the predictions.

## Model Results and Visualizations

The model coefficients and summary statistics are presented in Table @tbl-modelresults. These results provide insights into how different factors—such as poll quality and sample size—affect the probability of winning.

@@

# Load the election model
election_model <- readRDS(file = here::here("models/election_model.rds"))
```
```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Model results for predicting the winner of the election"
#| warning: false

modelsummary::modelsummary(
  list(
    "Election Model" = election_model
  ),
  statistic = "mad",
  fmt = 2
)
```
